Tissue Enrichment Analysis (TEA)

### 2016-04-06 tests/hgf_benchmarking/doc/tea_main_text.tex
writing!!!

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
% \RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
% \usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{lmodern}% http://ctan.org/pkg/lm
\usepackage{subcaption} %for subfigures
\usepackage{graphicx} %graphics support to specify path
\usepackage{csvsimple} %insert tables using .csv files
\usepackage{caption}
\usepackage{adjustbox}

@@

\title{Tissue Enrichment Analysis: TEA}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been explosive development in tools
for measuring gene expression. These tools can identify thousands of altered 
genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging. One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
with controlled relations between terms) with genes and to look for enrichment of specific terms. Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomy information.
\parttitle{Results} %if any 
We have developed an  tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets. We applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to identify enriched tissues and generated a website GUI. To cut down on verbosity, we have come up with three straightforward filtering criteria that slim the ontology by almost tenfold. One filter removes all terms with fewer than an arbitrary number of annotating genes; the second filter removes all sister terms that have identical gene annotations; and the third filter removes all parents whose daughters all survived the previous two filters. We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA) can be found at www.WormBase.org/tea, and can be downloaded using Python's standard pip installer. It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results. Our tool can discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene. For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}. The \emph{C.~elegans} database, WormBase~\cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters. Here we provide a new framework that analyses a user-input list for enrichment of specific tissues. Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

@@

 To prevent our tool from suffering from verbosity problems, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has never been performed in an algorithmic fashion for a biomedical ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.

\section*{Results}
\subsection*{Generating a Useful Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. We also wanted to avoid
testing tissues at levels where redundancy becomes problematic. For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing. However, many left/right neuronal
pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them. We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. We refer to such sisters as `redundant'. To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}). Our similarity metric can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score. We referred to these two scoring criteria as `avg' and `any' respectively.

\subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
Another problem arises from the fact that the tissue ontology is scarcely populated at this point in time. Many nodes have 0-10 annotations, which we consider too few to accurately test. To solve this issue, we implemented a
straightforward trimming algorithm. For a given terminal node, we test whether the node has more than a threshold number of annotations. If it does not, the node is removed. The next node in the branch is tested and removed recursively until a node which satisfies the condition is found. At that point, no more nodes can be removed from that branch. This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}). In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
By itself, each of these filters can reduce the number of nodes employed for analysis. These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter. If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering will result in greater numbers of nodes. We always applied the ceiling at the end. For validation (see below) we made a number of different dictionaries. The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).

\subsection*{Tissue enrichment testing via a hypergeometric model}
Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model. Briefly, the hypergeometric model assumes the existence of an urn with a pre-determined number of balls inside it. The balls can be painted one of several colors. The hypergeometric model provides an answer to the question: If an individual removes $N$ balls, what is the probability of observing $n_i$ balls of color $i$, if the balls are selected without replacement? Mathematically, this is expressed as: 
\begin{eqnarray}\label{hypergeometric}
	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}
\end{eqnarray}

Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number  of draws, $m_i$ is tissue $i$ and $M$ is the total number of balls in the urn. In our specific case, $M_i$ is equal to the total number of annotations in our dictionary. $N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary. The remaining genes are then associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors. Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list. Due to the discrete nature of the hypergeometric distribution, this algorithm can generate artifacts when the list is small. To avoid spurious results, a tissue is never considered significant if there are no annotations for it in the user-provided list.

Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}. Genes that have a q-value less than a given alpha are considered significant. Our default setting is to set alpha at 0.1.

Users input a gene-list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WBIDs. The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits). Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. Our software is implemented in an easy to use GUI within WormBase (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique WBbt ID.

\subsection*{Validation of the algorithm and parameter selection}
We wanted to select a dictionary which included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues. On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations. To our knowledge there is no good method for assessing false-positive or false-negative results for
annotations.
% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011,
Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase. Some of these studies have
been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any'
thresholding criteria for the latter (see Table~\ref{tab:DictionarySpecs}). For these dictionaries, the number of tissues tested ranged from 21 to 460. The number of tissues was inversely correlated to the minimum annotation, as
expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1). Next, we analyzed all 30 datasets using each dictionary. Because of the large number of results, instead of analyzing each set of terms individually, we pooled all results for a given dictionary into histograms. When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary. We also noticed that the `any' thresholding method resulted in tighter histograms with fa mode closer to 0. For this reason, we chose the `any' method for dictionary generation. The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term
specificity. Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
% TODO add WBbt codes!
Most results were highly comparable and in line with what was expected. For some sets, all dictionaries seemed to perform well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}). On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
%Differences between dictionaries
We queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary suggested, probably correctly, that the E lineage was heavily enriched in this set.
%Bad stuff.
Not all queries worked equally well. For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms. We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
%Dictionary intra-agreement
We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues. We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}. Overall, the tool seems to have good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:intraagree}). However, most high-scoring terms were preserved between gene sets. The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement. All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95
redundancy threshold using the `any' criterion. seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of the biology underlying these datasets. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (for an example, see~\ref{fig:Dcon}\ignore{,~\ref{fig:Hsp},\ref{fig:Smar}}). A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. One bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

Chikina \emph{et al}~\cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies. SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data becomes more available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for some genome-wide studies. Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes. To our knowledge, there are no other tissue ontology enrichment tools in \emph{C.~elegans}, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
% \RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
% \usepackage[utf8]{inputenc} %unicode support
% \usepackage[applemac]{inputenc} %applemac support if unicode package fails
% \usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{lmodern}% http://ctan.org/pkg/lm
\usepackage{subcaption} %for subfigures
\usepackage{graphicx} %graphics support to specify path
\usepackage{csvsimple} %insert tables using .csv files
\usepackage{caption}
\usepackage{adjustbox}

@@

\title{Tissue Enrichment Analysis (TEA)}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been explosive development in tools
for measuring gene expression. These tools can identify thousands of altered 
genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging. One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
with controlled relations between terms) with genes and to look for enrichment of specific terms. Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomy information.
\parttitle{Results} %if any 
We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets. We applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to identify enriched tissues and generated a website GUI. To cut down on verbosity, we have come up with three straightforward filtering criteria that slim the ontology by almost tenfold. One filter removes all terms with fewer than an arbitrary number of annotating genes; the second filter removes all sister terms that have identical gene annotations; and the third filter removes all parents whose daughters all survived the previous two filters. We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA) can be found at \url{www.wormbase.org/tea}, and can be downloaded using Python's standard pip installer. It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results. Our tool can discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene. For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}. The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from GFP-reporters. Here we provide a new framework that analyses a user-input list for enrichment of specific tissues. Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

@@

 To prevent our tool from suffering from verbosity problems, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

%Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.

\section*{Results}
\subsection*{Generating a Useful Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic. For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing. However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them. We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. We refer to such sisters as `redundant'. To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}). Our similarity metric can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score. We referred to these two scoring criteria as `avg' and `any' respectively.

\subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
Another problem arises from the fact that the tissue ontology is still scarcely populated. Many nodes have 0-10 annotations, which we consider too few to accurately test. To solve this issue, we implemented a straightforward trimming strategy. For a given terminal node, we test whether the node has more than a threshold number of annotations. If it does not, the node is removed. The next node in the branch is tested and removed recursively until a node which satisfies the condition is found. At that point, no more nodes can be removed from that branch. This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}). In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
By itself, each of these filters can reduce the number of nodes employed for analysis. These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is commutative with neither the trimming nor the redundancy filter. If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering will result in greater numbers of nodes. We always applied the ceiling at the end. For validation (see below) we made a number of different dictionaries. The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).

\subsection*{Tissue enrichment testing via a hypergeometric model}
Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model. Briefly, the hypergeometric model assumes the existence of an urn with a pre-determined number of balls inside it. The balls can be painted one of several colors. The hypergeometric model provides an answer to the question: If an individual removes $N$ balls, what is the probability of observing $n_i$ balls of color $i$, if the balls are selected without replacement? Mathematically, this is expressed as: 
\begin{eqnarray}\label{hypergeometric}
	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}\text{.}
\end{eqnarray}

Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number  of draws, $m_i$ is tissue $i$ and $M$ is the total number of balls in the urn. In our specific case, $M_i$ is equal to the total number of annotations in our dictionary. $N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary. The remaining genes are associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors. Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list. Due to the discrete nature of the hypergeometric distribution, this algorithm can generate artifacts when the list is small. To avoid spurious results, a tissue is never considered significant if there are no annotations for it in the user-provided list.

Once the p-values for each term have been calculated, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}. FDR corrected p-values are called q-values. Genes that have a q-value less than a given alpha are considered significant. Our default setting is to set alpha at 0.1.

Users input a gene-list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WormBase gene IDs (WBGene IDs). The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits). Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. Our software is implemented in an easy to use GUI within WormBase (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).

\subsection*{Validation of the algorithm and parameter selection}
We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. As expected, larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
%The phrase below doesn't seem 
%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}). The number of tissues was inversely correlated to the minimum annotation cutoff, as expected, and was largely insensitive to the similarity threshold in the range we explored (0.9-1). Next, we analyzed all 30 datasets using each dictionary. Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset. When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary. We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0. For this reason, we chose the `any' method for dictionary generation. The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity. Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity. 
%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
% TODO add WBbt codes!
Most results were comparable and in line with expectations. For some sets, all dictionaries performed well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron-related terms %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays 
regardless of the dictionary used (see Table~\ref{tab:interagree}). On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells---at the 33-cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as being five-fold enriched, and at the 25 gene-cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
%Differences between dictionaries
We also queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched in this set.
%Bad stuff.
Not all queries worked equally well. For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine related terms in any dictionary, but they were enriched for pharynx- and hypodermis-related terms. We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.

%Dictionary intra-agreement
We assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues. We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD enriched sets~\cite{Spencer2011, Smith2010}; two  GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}. Overall, the tool seems to have good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}). However, most high-scoring terms were preserved between gene sets. %The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement. 
All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion. seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (for an example, see~\ref{fig:Dcon}\ignore{,~\ref{fig:Hsp},\ref{fig:Smar}}). A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. One bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

Chikina \emph{et al}~\cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies. SVM classifiers are powerful tools, but they require continuous retraining as more tissue expression data becomes available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies. Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes. To our knowledge, there are no other tissue ontology enrichment tools in \emph{C.~elegans}, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.

We have tried hard to benchmark our tool well. However, our analysis suffers from the drawback that is very hard to %identify good negative controls
  to identify spurious term enrichment. % Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results. For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33. Is this an error, or does this hint at new biology? 
 Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set. For example, we know that expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}. As time passes, we are certain the accuracy and power of this tool will improve thanks to the efforts of the combined worm community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
